\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2015-01-15 tex/introduction.tex
reduce number of references

%!TEX root = ../paper.tex
The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to derive a system of differential equations to model a particular biological network, attempt to fit that model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. All of these models utilize essentially equivalent mathematical structures \ref{fig:biomodelexamples} \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. Developing unified mathematical descriptions of all of these is one of the paramount goals of systems biology.

Recent work has demonstrated that as a result of the existence of largely insensitive directions within the parameter space for such models, the approach outlined above often allows for a large variety of models to fit equivalent data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In addition, there is often uncertainty about the very structure of such networks. In this context, it is crucial to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}. This can be achieved using a method common in statistical physics involving the consideration of an ensemble of systems that, in comparison to one another, appear to have components that are randomly interlinked.

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is, however, not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the networks represented in \ref{fig:biomodelexamples}. May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014,Majumdar2014}, which implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

@@

Here we determine the relationship between network hierarchy, a topological property, and the probability of \emph{robustness}, a dynamical property. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of gene-regulatory networks \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}. Intuitively, robustness is the probability of stability to potentially large perturbations of the parameters for systems which have already been determined to be stable in the sense of linear stability analysis \cite{Davis1962}. The maximally hierarchical network is considered to be the graph associated to the total ordering, which is the complete directed acyclic graph for a given system size \cite{Cormen2009}. Here, hierarchy is then defined as the edit distance from this maximally hierarchical network \cite{Axenovich2011}. We demonstrate that systems with the most hierarchical network topology according to this definition exhibit maximal robustness and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology. Moreover these results hold for networks of arbitrary size and independent of the probability distribution from which the strengths of interaction are sampled.

>>>>

%!TEX root = ../paper.tex
The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to derive a system of differential equations to model a particular biological network, attempt to fit that model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. All of these models utilize essentially equivalent mathematical structures \ref{fig:biomodelexamples} \cite{RossCr2003,Palsson2011a,Sauro2012}. Developing unified mathematical descriptions of all of these is one of the paramount goals of systems biology.
% \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}

Recent work has demonstrated that as a result of the existence of largely insensitive directions within the parameter space for such models, the approach outlined above often allows for a large variety of models to fit equivalent data \cite{Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In addition, there is often uncertainty about the very structure of such networks. In this context, it is crucial to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}. This can be achieved using a method common in statistical physics involving the consideration of an ensemble of systems that, in comparison to one another, appear to have components that are randomly interlinked.
% \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}

Investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is, however, not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the networks represented in \ref{fig:biomodelexamples}. May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014,Majumdar2014}, which implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

@@

Here we determine the relationship between network hierarchy, a topological property, and the probability of \emph{robustness}, a dynamical property. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of gene-regulatory networks \cite{VanNimwegen1999,Siegal2002,Wagner2013}. Intuitively, robustness is the probability of stability to potentially large perturbations of the parameters for systems which have already been determined to be stable in the sense of linear stability analysis \cite{Davis1962}. The maximally hierarchical network is considered to be the graph associated to the total ordering, which is the complete directed acyclic graph for a given system size \cite{Cormen2009}. Here, hierarchy is then defined as the edit distance from this maximally hierarchical network \cite{Axenovich2011}. We demonstrate that systems with the most hierarchical network topology according to this definition exhibit maximal robustness and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology. Moreover these results hold for networks of arbitrary size and independent of the probability distribution from which the strengths of interaction are sampled.
% \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}

### 2015-01-15 tex/methods.tex
minor edits
reduce number of references

%!TEX root = ../paper.tex
\subsection{Monte carlo integration}
If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
\begin{widetext}
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
\end{widetext}
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation.

>>>>

%!TEX root = ../paper.tex
\subsection{Monte carlo integration}
If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
% \begin{widetext}
\begin{align*}
&\mathrm{var}(\theta | \mathcal{D}) =\\
&\frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
\end{align*}
% \end{widetext}
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

### 2015-01-15 tex/overview.tex
reduce number of references

The network architecture can be represented in terms of an adjacency matrix and further abstracted by mapping the interaction graph to the network of strongly connected components (SCCs, see Supplementary Information) (\reffigscc). This map from the interaction graph of a network, referred to as $\hier$, has a collection of symmetries shown in \reffighiertransformations. These three symmetries represent transformations that can be performed on the interaction graph that do not change the network of SCCs to which it is associated \reffigscc. Two of these three are also symmetries with respect to dynamical robustness. \ref{fig:robustnesssymmetries} shows an example of these symmetries applied to a specific interaction graph.

>>>>

The network architecture can be represented in terms of an adjacency matrix and further abstracted by mapping the interaction graph to the network of strongly connected components (SCCs, see \refsupp{}) (\reffigscc). This map from the interaction graph of a network, referred to as $\hier$, has a collection of symmetries shown in \reffighiertransformations. These three symmetries represent transformations that can be performed on the interaction graph that do not change the network of SCCs to which it is associated \reffigscc. Two of these three are also symmetries with respect to dynamical robustness. \ref{fig:robustnesssymmetries} shows an example of these symmetries applied to a specific interaction graph.

### 2015-01-15 tex/supplement.tex
reduce number of references

The network architecture can be represented in terms of an adjacency matrix and further abstracted by mapping the interaction graph to the network of strongly connected components (SCCs, see Supplementary Information) (\reffigscc). This map from the interaction graph of a biological network, referred to as $\hier$, has a collection of symmetries shown in \reffighiertransformations. These three symmetries represent transformations that can be performed on the interaction graph that do not change the network of SCCs to which it is associated \reffigscc. Two of these three are also symmetries with respect to dynamical robustness. \ref{fig:robustnesssymmetries} shows an example of these symmetries applied to a specific interaction graph.

>>>>

The network architecture can be represented in terms of an adjacency matrix and further abstracted by mapping the interaction graph to the network of strongly connected components (SCCs, see \refsupp{}) (\reffigscc). This map from the interaction graph of a biological network, referred to as $\hier$, has a collection of symmetries shown in \reffighiertransformations. These three symmetries represent transformations that can be performed on the interaction graph that do not change the network of SCCs to which it is associated \reffigscc. Two of these three are also symmetries with respect to dynamical robustness. \ref{fig:robustnesssymmetries} shows an example of these symmetries applied to a specific interaction graph.
